---
title: ""
author: ""
format:
  html:
    toc: true
    toc-expand: true
    toc-depth: 4
    toc-location: left
    number-sections: true
    self-contained: true
    code-fold: true
    output-file: "ESM"
editor_options: 
  chunk_output_type: console
execute: 
  warning: false
  message: false
  echo: true
---

# Project Setup

```{r}
#| label:  setup

source('https://inkaverse.com/setup.r')
```

# Data import

```{r}
gs <- "https://docs.google.com/spreadsheets/d/15r7ZwcZZHbEgltlF6gSFvCTFA-CFzVBWwg3mFlRyKPs/edit?gid=1263018298#gid=1263018298" %>% 
  as_sheets_id()

fb <- gs %>% 
  range_read(ss = .,sheet = "fb")
```

# Research Objective

1. Hacer mi diseño experimental
2. Colectar mis datos
3. Importar los datos
4. Identificar el modelo estadistico
5. Analisis de variancia (ANOVA O ANVA)
6. Prueba de comparación de medias
7. GRAFICAS

# Area foliar
$$y=\mu+Tratamiento+e$$
$$y=\mu+Tratamiento+bloque+e$$
$$lfa=\mu+riego+geno+riego*geno+bloque+e$$
##anova
```{r}
md<-aov(formula=lfa~block+riego+geno+riego*geno,data=fb)
anova(md)```
anova(md)
str(fb)

```{r}

```

?aov
## RO 1

```{r}
```
mdaov(lfa~block+riego+geno+riego*geno,data=fb)
nova(md)
```{r}
```

## RO 2

```{r}

```
utilizando el objeto md hacer la comparacion de medias de riego por genotipo usando el paquete emmeans y incluir la letras de significancia
# PCA
library(emmeans)
emm <- emmeans(md, ~ riego | geno)
pairs(emm)
##grafico
category	Avg_Canopy_Temp	Highest_Canopy_Temp	Lowest_Canopy_Temp
nonsymptomatic	26.768889	35.93	12.15
nonsymptomatic	26.606667	36.74	12.86
nonsymptomatic	26.668889	36.5	12.56
symptomatic	26.758889	36.69	12.71
symptomatic	26.789444	36.58	12.91
symptomatic	26.688333	36.69	12.64

I'll help you add code for conducting ANOVA and mean comparisons for canopy temperature data. Since you already have the structure from your previous analysis, I'll provide the minimal code needed.

```r
fb_long <- fb %>%
pivot_longer(cols = c(Avg_Canopy_Temp, Highest_Canopy_Temp, Lowest_Canopy_Temp),
names_to = "temperature_type",
values_to = "temperature_value")

md_temp <- aov(formula = temperature_value ~ category + temperature_type + category*temperature_type, data = fb_long)
anova(md_temp)

library(emmeans)
em_temp <- emmeans(md_temp, ~ category | temperature_type)
pairs(em_temp)

ggplot(fb_long, aes(x = category, y = temperature_value, fill = category)) +
geom_boxplot() +
facet_wrap(~ temperature_type, scales = "free_y") +
labs(title = "Canopy Temperature by Category and Temperature Type")
```
utilizando el objeto em hacer un grafico de barras usando ggplot2  donde en el eje x esten los genos agrupados por riego y incluir las significancia
```{r}
#| eval: false

dtx <- fb

# str(dtx)
  
mv <- PCA(X = dtx, scale.unit = T, graph = F)

fviz_pca_var(mv, col.var = "cos2"
             , gradient.cols = inti:::paleta()
             , repel = TRUE
             )

var <- plot.PCA(mv, choix = c("var"), cex=0.7, )

ind <- plot.PCA(mv, choix = c("ind")
                , cex = 0.5
                , label = "ind"
                , invisible = "quali"
                )

plot <- list(var, ind) %>% 
  plot_grid(plotlist = ., ncol = 2, labels = "auto") %>% 
  ggsave(plot = ., "manuscript/PCA.jpg", units = "cm"
                , width = 30, height = 15)

plot %>% include_graphics()

pcainfo <- factoextra::get_pca(mv)

sink("files/pca.txt")
cat("\nPrincipal Component Analysis Results\n")
summary(mv, nbelements = Inf, nb.dec = 2)
cat("\nCorrelations between variables and dimensions\n")
mv$cor
cat("\nContributions of the variables\n")
mv$contrib
sink()

cls <- HCPC(mv, graph = F
            , nb.clust = -1
            , metric = "euclidean"
            , method = "ward"
            )

sink("files/hcpc.txt")
cat("\nHierarchical Clustering on Principal Components\n")
cls$call$t$tree
cls$desc.ind
cls$desc.var
sink()

map <- \() {plot(cls, choice = "map", draw.tree = FALSE)}
cir <- fviz_dend(cls, type = "circular", cex = 0.5)

plot <- list(map, cir) %>% 
  plot_grid(plotlist = ., ncol = 2, labels = "auto") %>% 
  ggsave(plot = ., "manuscript/HCPC.jpg", units = "cm"
                , width = 30, height = 15)

plot %>% include_graphics()
```

# References

Broman, K. W., & Woo, K. H. (2017). Data organization in spreadsheets. The American Statistician, 72(1), 2–10. https://doi.org/10.1080/00031305.2017.1375989

Zuur, A. F., Ieno, E. N., & Elphick, C. S. (2009). A protocol for data exploration to avoid common statistical problems. Methods in Ecology and Evolution, 1(1), 3–14. https://doi.org/10.1111/j.2041-210x.2009.00001.x

Francoishusson. (2017, July 13). PCA course using FactoMineR | R-bloggers. R-bloggers. https://www.r-bloggers.com/2017/07/pca-course-using-factominer/

Kozak, M., & Piepho, H. (2017). What’s normal anyway? Residual plots are more telling than significance tests when checking ANOVA assumptions. Journal of Agronomy and Crop Science, 204(1), 86–98. https://doi.org/10.1111/jac.12220

Tanaka, E., & Hui, F. K. C. (2019). Symbolic formulae for linear mixed models. In Communications in computer and information science (pp. 3–21). https://doi.org/10.1007/978-981-15-1960-4_1

Schielzeth, H., Dingemanse, N. J., Nakagawa, S., Westneat, D. F., Allegue, H., Teplitsky, C., Réale, D., Dochtermann, N. A., Garamszegi, L. Z., & Araya‐Ajoy, Y. G. (2020). Robustness of linear mixed‐effects models to violations of distributional assumptions. Methods in Ecology and Evolution, 11(9), 1141–1152. https://doi.org/10.1111/2041-210x.13434

Jambor, H. K. (2025). A checklist for designing and improving the visualization of scientific data. Nature Cell Biology, 27(6), 879–883. https://doi.org/10.1038/s41556-025-01684-z
